\chapter{Astrophysical Processes}
\minitoc
\pagebreak
\section{Quantifying radiation}
\subsection{Flux \& intensity}
Recalling from \textit{Introduction to Astrophysics \& Cosmology} we define the energy flux density (or simply flux), $F$ as the power per unit surface area.
%
$$ \diff E = F \diff A \diff t $$
%
Note that flux is dependant on the orientation of the unit surface with respect to the Poynting vector of the radiation field.
 Flux has units of J/s/m$^2$.
 Often it is more useful to use the \emph{monochromatic} energy flux density, $F_\nu$ where
%
\begin{equation}
	\diff E = F_\nu \diff A \diff t \diff \nu
	\label{eq:AP:def_flux}
\end{equation}
%
This gives the energy flux per unit of bandwidth, (frequency interval of radiation) which has units J/s/m$^2$/Hz.
\par 
In order to account for the differing directions of photons, the flux is further generalised by defining the specific monochromatic intensity, $I_\nu$.
 This quantifies the power per unit frequency from a range of directions covering a solid angle element $\diff \Omega$ given by
%
\begin{equation}
	\diff E = I_\nu \diff A dt \diff \nu \diff \Omega.
	\label{eq:AP:def_intensity}
\end{equation}
%
This then has units J/s/m$^2$/Hz/sr.
 Unlike flux, the intensity is constant along the direction of propagation of photons (in free space).
\par
From the respective definitions, we can relate flux and intensity by
%
\begin{align*}
	F_\nu \diff A &= \int \mathbf{I_\nu} \dotp \mathbf{\diff A} \diff \Omega
	\linedrop
	F_\nu \diff A &= \int I_\nu \cos(\theta) \diff A \diff \Omega
	\linedrop
	\therefore F_\nu &= \int I_\nu \cos(\theta) \diff \Omega
\end{align*}
%

\subsubsection{Solid angles: an aside}
The solid angle element $\diff \Omega$ is defined
%
$$ \diff \Omega = sin(\theta) \diff \theta \diff \phi , $$
%
hence when integrated over a complete sphere, $\Omega = 4\pi$~sr.
 This corresponds to the area of a sphere with unit radius.
 Considering the fraction of the total surface area, $A$ of a sphere with radius $R$, we find that
%
\begin{align}
	\frac{A}{4\pi R^2} &= \frac{\Omega}{4\pi}
	\linedrop
	\Omega &= \frac{A}{R^2}.
	\label{eq:AP:solid_angle}
\end{align}
%
As asserted previously, the solid angle corresponds to the area on a sphere with unit radius.
%
\subsubsection{Surface brightness}
Consider an object which a circular region in the sky with radius $\theta_c$.
 We choose to position this object such that it is located at the pole of the coordinate system.
 This is done for convenience such that the setup is rotationally symmetric in $\phi$.
 We can then calculate the flux incident on an aperture.
%
\begin{align*}
	F_\nu &= 2\pi I\nu \int_{0}^{\theta_c}\diff\theta \sin(\theta) \cos(\theta)
	\linedrop
	F_\nu &= \pi I_\nu \sin^2(\theta_c)
\end{align*}
%
For small $\theta_c$, $\sin(\theta_c) \simeq \tan(\theta_c) = R/r$ where $R$ is the object radius and $r$ is the object-aperture distance, hence
%
\begin{equation}
	F_\nu = \pi I_\nu  \left( \frac{R}{r} \right) ^2 \propto \frac{1}{r^2}.
	\label{eq:AP:flux_intensity_relation}
\end{equation}
%
For an isotropically emitting object (such as a star) the flux can be related to the luminosity (total power generated) by
%
$$ F = \frac{L}{4\pi r^2}. $$
%
%
%
\section{Radiative transfer}
\subsection{Free space}
We want to know how intensity, $I_\nu$ changes along the path of the ray.
 Consider two areas placed parallel to each other with a line length $R$ connecting their centres.
 If the areas are small relative to their separation, then all rays that pass through \emph{both} surfaces are effectively normal to the surfaces.
 From energy conservation,
 %
\begin{align*}
	\diff E_1 &= \diff E_2
	\linedrop
	I_{\nu ,1}\diff A_1 \diff t \diff \nu_1 \diff \Omega_1 &= I_{\nu ,2}\diff A_2 \diff t \diff \nu_2 \diff \Omega_2 ,	
\end{align*}
%
Since, in free space, the frequency remains constant and using equation \ref{eq:AP:solid_angle}:
%
\begin{align*}
		I_{\nu ,1}\diff A_1 \frac{\diff A_2}{R^2} &= I_{\nu ,2}\diff A_1 \frac{\diff A_1}{R^2}
		\linedrop
		I_{\nu ,1} &= I_{\nu ,2}.
\end{align*}
%
Given that the two surfaces have arbitrary size, we can conclude that in general, intensity is conserved along ray paths in free space.
 This is \emph{not} the case for flux.
 Intensity is conserved since it is defined as `per unit solid angle'.
 Hence, for a distance $s$ along the path of a ray,
%
$$ \frac{\diff I_\nu}{\diff s} = 0. $$
%
This is the equation of radiative transfer in free space.
%
%
\subsection{In a medium}
We now need  to consider emission generated along the line of sight as well as absorption by the medium.
 This can be quantified by the equation of radiative transfer in a medium,
%
\begin{equation}
	\frac{\diff I_\nu}{\diff s} = j_\nu - \alpha_\nu I_\nu,
	\label{eq:AP:radiative_transfer}
\end{equation}
%
where $j_\nu$ is the \emph{monochromatic emission coefficient} with units J/s/m$^3$/Hz/sr, and $\alpha_\nu$ is the \emph{monochromatic absorption constant} with units m$^{-1}$.
 It is simple to realise that doubling the intensity passing through a medium would in turn double the intensity absorbed per metre.
 Note that in free space, $j_\nu = 0$ and $\alpha_\nu = 0$.
\par
\subsubsection{Emission coefficient}
From the definition of intensity given by equation \ref{eq:AP:def_intensity}, but changing $\diff E$ to $\delta_E$,
%
\begin{align}
	\frac{\diff \delta_E}{\diff s} &= \frac{\diff I_\nu}{\diff s} \diff A \diff t \diff \nu \diff \Omega
	\linedrop
	\frac{\diff \delta_E}{\diff s} &= j_\nu \diff A \diff t \diff \nu \diff \Omega
	\linedrop
	\therefore \diff \delta_E &= j_\nu \diff V \diff t \diff \nu \diff \Omega.
	\label{eq:AP:emission_coef}
\end{align}
%
From equation \ref{eq:AP:emission_coef} we can see that $j_\nu$ corresponds to the amount of energy radiated per unit volume\footnote{Note that $\diff V = \diff A \diff s$.}, per unit time, per unit solid angle for a frequency interval.
\par 
\subsubsection{Absorption constant}
By using similar treatment for the absorption, we find that
%
\begin{align}
		\frac{\diff \delta_E}{\diff s} &= -\alpha_\nu I_\nu  \diff A \diff t \diff \nu \diff \Omega
		\linedrop
		\frac{\diff \delta_E}{\diff s} &= -\alpha_\nu \delta_E.
		\label{eq:AP:absorption_const}
\end{align}
%
Equation \ref{eq:AP:absorption_const} implies that $\alpha_\nu \diff s$ represents the fraction of energy that disappears from the radiation field after travelling a length $\diff s$ through the medium.
\par  
\subsubsection{Opacity}
Following from the definition in equation \ref{eq:AP:absorption_const}, we now consider a volume of absorbing particles with number density $n$ and absorption cross section $\sigma_\nu$.
 Since each particle obscures an area $\sigma_\nu$, the total area blocked is $n\sigma_\nu\diff V$.
 Also, the fractional area (through which photons flow) that is blocked by absorbers is
$$ n \sigma_\nu \frac{\diff V}{\diff A} = n \diff s \sigma_\nu, $$
hence
% 
\begin{align*}
	\alpha_\nu &= n \sigma_\nu	
	\linedrop
	\alpha_\nu &= \rho \kappa_\nu.
\end{align*}
%
The \emph{opacity coefficient} $\kappa_\nu$ is defined such that $\sigma_\nu \propto \kappa_\nu$. Also, $\kappa_\nu$ has units m$^2$/kg.
\par 
\subsubsection{Optical depth}
A useful quantity to define is optical depth $\tau_\nu$, which is a re-scaled distance co-ordinate defined by
$$ \diff \tau_\nu \equiv \alpha_\nu \diff s, $$
which provides a more relevant unit to express distance in, rather than metres.
 The total optical depth can then be defined as
$$ \tau_\nu (s) = \int_{s_0}^{s} \alpha_\nu \diff s^\prime. $$
The optical depth becomes particularly useful in the case $j_\nu = 0$.
 In this situation, the equation of radiative transfer (see equation \ref{eq:AP:radiative_transfer}) becomes
$$ \diff I_\nu = - I_\nu \diff \tau_\nu. $$
This differential equation is easy to solve and we find that
%
\begin{equation}
I_\nu (\tau_\nu) = I_{\nu,0} e^{\tau_\nu}.
\end{equation}
%
\par 
\subsubsection{Source function}
Another useful quantity to define is the source function which follows a similar derivation. %REF Draine 7.4
This, along with the definition for optical depth can be used to further generalise the equation of radiative transfer (\ref{eq:AP:radiative_transfer}).
 The source function, $S_\nu$ is defined as
%
\begin{equation}
	S_\nu \equiv \frac{j_\nu}{\alpha_\nu}.
\end{equation}
%
The radiative transfer equation can now be defined using both $S_\nu$ and $\tau_\nu$ as
%
\begin{align*}
	\diff I_\nu &= \frac{j_\nu}{\alpha_\nu} \diff \tau_\nu - I_\nu \diff \tau_\nu
	\linedrop
	\diff I_\nu &= S_\nu \diff \tau_\nu - I_\nu \diff \tau_\nu.
\end{align*}
%
We can then integrate this to obtain the solution by introducing an integration factor $e^{\tau_\nu}$.
%
\begin{align*}
	\frac{\diff I_\nu}{\diff \tau_\nu} + I_\nu &= S_\nu
	\linedrop
	e^{\tau_\nu} \left( \frac{\diff I_\nu}{\diff \tau_\nu} + I_\nu \right) &= e^{\tau_\nu} S_\nu
	\linedrop
	\text{product rule} \implies 
	\frac{\diff}{\diff \tau_\nu} \left( e^{\tau_\nu} I_\nu \right) &= e^{\tau_\nu} S_\nu
	\linedrop
	\diff \left( e^{\tau_\nu} I_\nu \right) &= e^{\tau_\nu} S_\nu \diff \tau_\nu
	\linedrop
	\int_{I_\nu (0)}^{e^{\tau_\nu} I_\nu} \diff (e^{\tau_\nu} I_\nu) &= \int_{0}^{\tau_\nu} e^{\tau^{\prime}_\nu} S_\nu \diff \tau^\prime_\nu
\end{align*}
%
Performing this integration, and then dividing by the introduced factor of $e^{\tau_\nu}$ gives the general solution to the equation of radiative transfer, shown in equation \ref{eq:AP:radiative_transfer_gen}
%
\begin{equation}
	I_\nu (\tau_\nu) = I_\nu (0) e^{-\tau_\nu} + \int_{0}^{\tau_\nu} e^{-(\tau_\nu - \tau^\prime_\nu)} S_\nu \diff \tau^\prime_\nu.
	\label{eq:AP:radiative_transfer_gen}
\end{equation}
%
As we have seen previously, radiation from a background source is attenuated with an exponential factor given by the optical depth.
 The second term in equation \ref{eq:AP:radiative_transfer_gen} is the intensity generated in the medium due to emission.
 It is apparent that any emission produced is also attenuated by the medium between the point of generation of the emission and the observer.
 For constant $S_\nu$, we find that
%
\begin{equation}
	I_\nu (\tau_\nu) = I_\nu (0) e^{-\tau_\nu} + S_\nu (1 - e^{-\tau_\nu}).
	\label{eq:AP:radiative_const_S}
\end{equation}
%
The limit $\tau_\nu \gg 1$ is the \emph{optically thick} limit, and $\tau_\nu \ll 1$ is the \emph{optically thin} limit.
%
%
%
\section{Black body radiation}
Radiation is referred to as `thermal' if the spectrum depends \emph{only} on the temperature of the emitting body.
 The spectrum of `non-thermal' radiation however, is affected by other parameters such as chemical composition, or magnetic field strength.
 A perfect absorber in local thermal equilibrium would radiate black body radiation.
 A commonly cited example would be photons escaping through a small hole in a box containing a ``photon gas'' in equilibrium with the box.
 For isotropic radiation, the spectrum is given by the \emph{Planck function} shown in equation \ref{eq:AP:planck_func}
%
\begin{equation}
	I_\nu \equiv B_\nu = \frac{2 h \nu ^3}{c^2} \frac{1}{\exp(\frac{h \nu}{k_B T}) - 1}.
	\label{eq:AP:planck_func}
\end{equation}
%
The Planck curve is monotonic meaning that for a larger $T$, all points on the curve are higher; the curves do not cross.
%
\subsubsection{Changing variables}
We can express the Planck function in terms of wavelength rather than frequency.
 In order to do this we must follow this procedure:
 consider a frequency interval $(\nu + \diff \nu)$ and it's corresponding wavelength interval $(\lambda + \diff \lambda)$.
 We require that the intensity should be the same when considering these intervals, i.e.\ $B_\nu \diff \nu \equiv B_\lambda \diff \lambda$ hence,
%
\begin{align}
	B_\lambda &= B_\nu \left| \frac{\diff\nu}{\diff\lambda} \right|				\nonumber
	\linedrop
	B_\lambda &= B_\nu \left| \frac{\diff (c/\lambda)}{\diff\lambda}\right|		\nonumber
	\linedrop
	B_\lambda &= B_\nu \frac{c}{\lambda^2}										\nonumber
	\linedrop
	B_\lambda &= \frac{c}{\lambda^2} \frac{2 h (c/\lambda)^3}{c^2} \frac{1}{\exp(\frac{h c}{k_B T \lambda}) - 1} \nonumber
	\linedrop
	B_\lambda &= \frac{2 h c^2}{\lambda^5} \frac{1}{\exp(\frac{h c}{k_B T \lambda}) - 1}
\end{align}
%
Note that we use the absolute value of the derivative since both $\diff\lambda$ and $\diff\nu$ were defined to be positive whereas the derivative is negative.            
%
\subsubsection{High and low frequency limits}
The long and short wavelength limits were determined empirically before the theoretical derivation of the Planck function.
\par 
The high frequency (short wavelength) limit where $h\nu \gg k_B T$ is known as the \emph{Wien} limit.
 The exponential term is much larger than one, hence
%
\begin{equation}
	B_\nu \simeq \frac{2 h \nu^3}{c^2} \exp\left(\frac{-h\nu}{k_BT}\right).
	\label{eq:AP:wien_limit}
\end{equation}
%
\par 
The low frequency (long wavelength) limit where $h\nu \ll k_B T$ is known as the \emph{Rayleigh-Jeans} limit.
 In this limit, the exponential term is expanded as $\exp \left( h\nu/k_BT \right) = 1 + h\nu/k_BT + ...$ Hence,
%
\begin{equation}
	B_\nu \simeq \frac{2 \nu^2 k_B}{c^2} T.
	\label{eq:AP:rayleigh_jeans_limit}
\end{equation}
%
In this limit, intensity is proportional to temperature.
 As a result, in radio astronomy it is common to express intensities as `brightness temperatures'.
%
%
\subsection{Kirchhoff's law}
Consider an optically thick cloud with a uniform temperature which emits thermal radiation.
 The radiative transfer equation (eq. \ref{eq:AP:radiative_const_S}) can be written as
%
\[ \diff I_\nu = S_\nu \diff \tau_\nu - B_\nu \diff \tau_\nu = 0, \]
%
since \( I_\nu = B_\nu \) everywhere inside the cloud.
 This implies that
%
\begin{equation}
	B_\nu = S_\nu \equiv \frac{j_\nu}{\alpha_\nu},
	\label{eq:AP:kirchhoff's_law}
\end{equation}
%
which is known as \emph{Kirchhoff's law}: the source function of thermal radiation is the Planck function.
 Note that in general, the observed intensity is a combination of the background and source function.
%
%
\subsection{Temperature measurements}
\subsubsection{Brightness temperature}
We define a brightness temperature \(T_b\) for an intensity measurement at a certain frequency such that
%
\begin{equation}
	I_\nu = B_\nu (T_b).
	\label{eq:AP:brightness_temp}
\end{equation}
%
This gives a temperature which corresponds to the Planck function which reproduces the flux observed in a particular frequency band.
 Only in the case that the source is producing black body radiation does the temperature of the source equal \(T_b\).
%
\subsubsection{Stefan Boltzmann law}
The bolometric flux at the surface of a black body is given by
%
\begin{equation}
	F_\text{surf} = \sigma_{SB} T^4,
	\label{eq:AP:stefan_boltzmann_law}
\end{equation}
%
where \( \sigma_{SB} \) is the Stefan-Boltzmann constant ( \( 5.67 \times 10^{-8} \) J/s/m\( ^2 \)/K\( ^4 \) ).
 This shows that a unit area of black body surface with uniform temperature radiates a power \( \sigma_{SB} T^4 \).
 Hence, a measured intensity (assuming the intensity is unaffected by the medium along the line of sight) is related to temperature by
%
\begin{equation}
	\pi I = \sigma_{SB} T_\text{eff}^4,
	\label{eq:AP:effective_temperature}
\end{equation}
%
where \( T_\text{eff} \) is the \textit{effective temperature} remembering that it is the temperature of a black body emitting the same observed flux.
 Note that the size of the object should be known before determining the effective temperature, in order to relate the measured flux to the intensity (see eq. \ref{eq:AP:flux_intensity_relation}).
%
\subsubsection{Wien displacement law}
The peak of the Planck function is temperature dependent and described by the Wien displacement law:
%
\begin{align}
	h \nu_\text{max} &= 2.8214 k_B T,
	\linedrop
	k_B T \lambda_\text{max} = 0.201 h c.
	\label{eq:AP:wien_displacement_law}
\end{align}
%
Wien's law implies that the colour of a black body emitter is defined by its temperature.
 The \emph{colour temperature} of an object can be determined by finding the ratio of intensities measured at two frequencies.
 As with the previous temperature definitions, the colour temperature only equals the actual temperature if the radiation is emitted by a black body.
%
%
%
\section{Fluid mechanics of shock waves}
In previous chapters we have covered the basic conservation laws in fluid mechanics.
 These include 


